---
output: html_document
params:
  chart_x_date: "2020-01-01"
  include_annex: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.height = 5,
	fig.width = 10,
	warning = FALSE,
	message = FALSE
)
library(here)
library(tidyverse)
library(tidyr)
library(lubridate)
library(ggplot2)
library(janitor)
library(scales)
library(purrr)

# functions --------------------------------------------------------------------
source(here("R", "functions.R"))
# source(here("R", "theme_fsr.R"))
# source(here("R", "palette_fsr.R"))

# data files--------------------------------------------------------------------
metadata <- read_rds(here("data", "tidy", "metadata.rds"))
mm23_month <- read_rds(here("data", "tidy", "mm23_month.rds"))
oecd <- readRDS(here("data", "tidy", "oecd.rds"))

# setup vars--------------------------------------------------------------------
release_date <- dmy(max(metadata$release_date))
next_release_date <- dmy(max(metadata$next_release))

# current data date
latest_date <- max(mm23_month$date)
latest_month <- month(latest_date, label = TRUE, abbr = FALSE)
latest_year <- year(latest_date)

# same date last year and month
last_year <- latest_date %m-% years(1)
last_month <- latest_date %m-% months(1)

# chart styling-----------------------------------------------------------------

# x axis limit for filters
chart_x_date <- params$chart_x_date

chart_text_size <- theme(axis.title.y=element_text(size=14),
                         axis.text.x = element_text(size=14),
                         axis.text.y = element_text(size=14),
                         legend.text=element_text(size=14),
                         plot.title = element_text(size = 18),
                         plot.subtitle = element_text(size = 14),
                         plot.caption = element_text(size = 9, face = "plain"))

# initial data wrangling--------------------------------------

# Contribution to annual rate
cdids <- c("L5HB","L5HA","L5HF","L5HE","L5HD","L5HC","L5HG","L5H6","L5H5","L5H9","L5H8","L5H7")

x <- map_df(cdids, get_series_metadata, df = metadata) %>% 
  mutate(title = str_to_sentence(tidy_series_name(title,start = "CPIH: Contribution to all items annual rate: ", end = "#"))) %>% 
  select(cdid, title)

cont_ann <- mm23_month %>% 
  filter(cdid %in% cdids & value != is.na(value) & date >= latest_date) %>% 
  mutate(rank = rank(-value),
         rank_text = scales::ordinal(rank(-value))) %>% 
  select(date, cdid, value, rank, rank_text) %>% 
  arrange(rank) %>% 
  left_join(x)




```

## Food price inflation[^1] summary: `r format(latest_date, "%B %Y")`

[^1]: By which, we refer to the ONS 'food and non -alcoholic beverage' price index.

#### ONS Data published at 0700 on `r format(release_date, "%d %B %Y")`.

##### Next data release: `r format(next_release_date, "%d %B %Y")`

### Annual inflation rate

```{r latest-headlines, results='asis'}

# Food is L55P
x <- get_val_month(mm23_month, "L55P", latest_date)
cat(paste0(
 " - **Food price inflation** in ", format(x$date, "%B %Y"), " was ", x$val, "%. This was ", eval_change(x$val, x$prev_mth_val), " from ", x$prev_mth_val, "% in ",
 format(x$prev_mth_date, "%B %Y"), ".\n"
)
)

x <- get_prev_high(mm23_month, cdid = "L55P", series_date = latest_date)
cat(paste0(
 "\n - The last time inflation was at this level was in ",  format(x$prev_high_mth, "%B %Y"),
 " when the inflation rate was ", x$prev_high_val, "%.\n"
)
)

# All items is L55O
x <- get_val_month(mm23_month, "L55O", latest_date)
cat(paste0(
 "\n - **Overall inflation** in ", format(x$date, "%B %Y"), " was ", x$val, "%. This was ", eval_change(x$val, x$prev_mth_val), " from ", x$prev_mth_val, "% in ",
 format(x$prev_mth_date, "%B %Y"), ".\n"
)
)

x <- get_prev_high(mm23_month, cdid = "L55O", series_date = latest_date)
cat(paste0(
 "\n - The last time inflation was at this level was in ",  format(x$prev_high_mth, "%B %Y"),
 " when the inflation rate was ", x$prev_high_val, "%.\n"
)
)


# Contribution bullet point
cat(paste0("\n - Food made the ", cont_ann[cont_ann$cdid == "L5H5",]$rank_text,
           " largest contribution to the change in rate.",
           "The two highest contribtutions were from ",
           cont_ann[cont_ann$rank == 1,]$title, " and ",
           cont_ann[cont_ann$rank == 2,]$title, ". \n"
           ))



```

### Monthly inflation[^2]

[^2]: Refers to changes in the price index for food/overall CPIH on a month-on-month basis. These numbers are normally more volatile than year-on-year inflation figures ask they focus on variations within a short timeframe.

```{r momdata, results='asis'}
# Monthly all items L59C, food L59D
cpih_all_mm <- "L59C"
cpih_food_mm <- "L59D"

x <- get_val_month(df = mm23_month, cdid = "L59D", series_date = latest_date)
cat(paste0(
  " - Month on month food inflation was ", x$val,
  "% in ", format(x$date, "%B %Y"),
  " compared to ", x$prev_mth_val,
  "% in ", format(x$prev_mth_date, "%B %Y"),
  ".\n"
))

x <- get_val_month(df = mm23_month, cdid = "L59C", series_date = latest_date)
cat(paste0(
  " - Month on month overall inflation was ", x$val,
  "% in ", format(x$date, "%B %Y"),
  " compared to ", x$prev_mth_val,
  "% in ", format(x$prev_mth_date, "%B %Y"),
  ".\n"
))


```

```{r foodcats, results='asis'}


tidy1 <- "CPIH MO?N?THLY RATE 01.1.[0-9] ?:? ?"
tidy2 <- " 2015=100"

cdids <- c("L59S","L59T","L59U","L59V","L59W","L59X","L59Y","L59Z","L5A2")

labels <- c("Bread and\n cereals",
  "Meat",
  "Fish",
  "Milk, cheese\n and eggs",
  "Oils\n and fats",
  "Fruit",
  "Vegetables\n including\n potatoes and\n tubers",
  "Sugar, jam,\n syrups,\n chocolate\n and\n confectionery",
  "Food\n products (nec)")

data <- mm23_month %>% 
  filter(cdid %in% cdids & value != is.na(value) & date == max(date)) %>% 
  select(cdid, value) %>% 
  # arrange(-value) %>% 
  slice_max(order_by = value, n = 3)

cat(paste0(
  " - The biggest monthly changes within food categories were for ",
  tidy_series_name(get_series_metadata(metadata, data$cdid[1])$title, tidy1, tidy2),
  " (",
  data$value[1],
  "%), ",
    tidy_series_name(get_series_metadata(metadata, data$cdid[2])$title, tidy1, tidy2),
  " (",
  data$value[2],
  "%) and ",
    tidy_series_name(get_series_metadata(metadata, data$cdid[3])$title, tidy1, tidy2),
  " (",
  data$value[3],
  "%). \n"
))

data <- mm23_month %>% 
  filter(cdid %in% cdids & value != is.na(value) & date == max(date)) %>% 
  select(cdid, value) %>% 
  # arrange(-value) %>% 
  slice_min(order_by = value, n = 3)

cat(paste0(
  " - The smallest monthly changes were ",
  tidy_series_name(get_series_metadata(metadata, data$cdid[1])$title, tidy1, tidy2),
  " (",
  data$value[1],
  "%), ",
    tidy_series_name(get_series_metadata(metadata, data$cdid[2])$title, tidy1, tidy2),
  " (",
  data$value[2],
  "%) and ",
    tidy_series_name(get_series_metadata(metadata, data$cdid[3])$title, tidy1, tidy2),
  " (",
  data$value[3],
  "%). "
))



```

> -   insert ONS commentary here
> -   insert ONS commentary here
> -   insert ONS commentary here
> -   insert ONS commentary here

```{r yoyseries}
# All items series: L55O, food series: L55P
cdids <- c("L55O", "L55P")

data <- mm23_month %>% 
        filter(cdid %in% cdids & value != is.na(value) & date >= chart_x_date)

ggplot(data) +
  geom_line(aes(x = date, y = value, colour = cdid), size = 1) +
  scale_colour_discrete(breaks = cdids, labels = c("Overall inflation", "Food inflation")) +
  scale_y_continuous(labels = label_percent(scale = 1),breaks = breaks_extended(10)) +
  scale_x_date(date_labels = "%b %Y", limits = c(ymd(chart_x_date), NA)) +
  labs(title = "General and Food inflation",
       subtitle = "year on year inflation",
       caption = "All items series: L55O, food series: L55P") +
  theme_minimal() +
  theme(axis.title = element_blank())
  # theme_fsr()  
  

```

```{r foodcats-chart}

cdids <- c("D7KR","D7KS","D7KT","D7KU","D7KV","D7KW","D7KX","D7KY","D7KZ")

labels <- c("Bread and\n cereals",
  "Meat",
  "Fish",
  "Milk, cheese\n and eggs",
  "Oils\n and fats",
  "Fruit",
  "Vegetables\n including\n potatoes and\n tubers",
  "Sugar, jam,\n syrups,\n chocolate\n and\n confectionery",
  "Food\n products (nec)")

data <- mm23_month %>% 
  filter(cdid %in% cdids & value != is.na(value) & date == max(date)) %>% 
  select(cdid, value)

ggplot(data) +
  geom_bar(aes(x = fct_reorder(cdid, -value), y = value),stat = "identity",  position = "dodge", fill = "red") +
  scale_x_discrete(breaks = cdids, labels = labels) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(title = "Food price category changes",
       subtitle = paste(format(latest_date, "%B %Y"), "compared to", format(last_month, "%B %Y")),
       caption = paste("series ids ", paste(cdids, collapse = " "))) +
  theme_minimal() +
  chart_text_size +
  theme(axis.title = element_blank(),
        panel.grid.minor = element_blank())


# paste(purrr::map2_chr(labels, cdids, paste), collapse = " ")

```

## Average retail prices of selected items

```{r average-rps}

cdids <- c('CZNT','CZOH','CZNW','VKYY','CZMY','CZNN')

# item
x <- purrr::map(cdids, get_series_metadata, df = metadata) 
col1 <- x %>% purrr::map(list("title")) %>% unlist


# latest prices
rp_compare <- function(seriesid, date1, date2){
  
text <- paste0(
  "£",
  get_val(mm23_month, seriesid, date1)/100,
  ", ",
  eval_change(get_val(mm23_month, seriesid, date1)/100, get_val(mm23_month, seriesid, date2)/100),
  " from £",
  get_val(mm23_month, seriesid, date2)/100,
  " in ",
  month(date2, label = TRUE, abbr = FALSE),
  " ",
  year(date2)
)

return(text)

}

col2 <- purrr::pmap(list( cdids, rep(latest_date, length(cdids)), rep(last_month, length(cdids)) ) , rp_compare) %>% unlist


# last year
col3 <- purrr::map2(cdids, last_year, get_val, df = mm23_month) %>% unlist %>% dollar(scale = 0.01, prefix = "£")



knitr::kable(data.frame(cdids, col1, col2, col3))

```

### International comparisons

```{r oecd, results='asis'}

countries <- c("GBR", "EU27_2020", "G-7", "FRA", "DEU", "USA", "JPN", "ITA", "CAN")
cat("#### Food price inflation rates in selected countries/regions:\n\n")

int <- oecd %>% 
  filter(subject == "FOOD" &
         # date == now &
         country_code %in% countries) %>% 
        group_by(country_code) %>% 
        slice_max(date) %>% 
        ungroup() %>% 
        mutate(country_name = fct_relevel(country_name, "United Kingdom"),
               value = paste0(sprintf("%1.1f", value),"%"),
               date = format(date, "%B %Y")) %>% 
        select(country_name, value, date) %>% 
        arrange(country_name)

knitr::kable(int)

oecd %>% 
  filter(subject == "FOOD" &
         # date == now &
         country_code %in% countries) %>% 
        group_by(country_code) %>% 
        slice_max(date) %>% 
        ungroup() %>% 
        mutate(isuk =  ifelse(country_code == "GBR", "Y", "N")) %>% 
        select(country_name, value, date, isuk) %>%
  ggplot() +
  geom_bar(aes(x = fct_reorder(country_name, -value), y = value, fill = isuk), stat = "identity") +
  scale_fill_manual(values = c("Y" = "darkred", "N" = "darkorange"), guide = "none") +
  labs(title = "Food price inflation") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        panel.grid.major.x = element_blank())

cat("\n\n#### Overall inflation rates in selected countries/regions:\n\n")

int <- oecd %>% 
  filter(subject == "TOT" &
         # date == now &
         country_code %in% countries) %>% 
        group_by(country_code) %>% 
        slice_max(date) %>% 
        ungroup() %>% 
        mutate(country_name = fct_relevel(country_name, "United Kingdom"),
               value = paste0(sprintf("%1.1f", value),"%"),
               date = format(date, "%B %Y")) %>% 
        select(country_name, value, date) %>% 
        arrange(country_name)

knitr::kable(int)

oecd %>% 
  filter(subject == "TOT" &
         # date == now &
         country_code %in% countries) %>% 
        group_by(country_code) %>% 
        slice_max(date) %>% 
        ungroup() %>% 
        mutate(isuk =  ifelse(country_code == "GBR", "Y", "N")) %>% 
        select(country_name, value, date, isuk) %>%
  ggplot() +
  geom_bar(aes(x = fct_reorder(country_name, -value), y = value, fill = isuk), stat = "identity") +
  scale_fill_manual(values = c("Y" = "darkred", "N" = "darkorange"), guide = "none") +
  labs(title = "Overall inflation") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        panel.grid.major.x = element_blank())


```

#### Next release: `r day(next_release_date)` `r month(next_release_date, label = TRUE, abbr = FALSE)` `r year(next_release_date)`

```{r annex, results='asis'}
if(params$include_annex == TRUE){
  cat("## Annex: useful background data\n\n")
  
 # Contribution data 
  cat("### Contribution to annual rate\n")
  
  print(knitr::kable(cont_ann))
  
  cat("\n\n")
  
  ggplot(cont_ann) +
  geom_bar(aes(x = fct_reorder(str_wrap(title,width = 8), -value), y = value, fill = as.factor(format(date, "%B %y"))),
           stat = "identity",
           position = position_dodge2(preserve = "single", reverse = TRUE)) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  scale_fill_viridis_d() +
  labs(title = paste("Contribution to annual rate,", format(latest_date, "%B %Y")),
       subtitle = "sub") +
  theme_minimal() + 
  chart_text_size +
  theme(axis.title = element_blank(),
        legend.title = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(size = 10),
        panel.grid.major.x = element_blank()) 

}
```

```{r momchart}

# cdids <- c("L59C", "L59D")
cdids <- c(cpih_all_mm, cpih_food_mm)

data <- mm23_month %>% 
  filter(cdid %in% cdids & value != is.na(value) & date >= chart_x_date)

ggplot(data) +
  geom_bar(aes(x = date, y = value, fill = cdid),
           stat = "identity",
           position = position_dodge(preserve = "single")) +
  # xlim(ymd("2021-01-01"), NA) +
  scale_fill_discrete(breaks = cdids, labels = c("All items", "Food")) +
  scale_x_date(date_labels = "%b %Y") +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 0.1), breaks = breaks_extended(5)) +
    labs(title = "General and Food inflation",
       subtitle = "Monthly inflation",
       caption = paste("All items series", cpih_all_mm, "food series", cpih_food_mm)) +
  theme_minimal() +
  chart_text_size +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        axis.title = element_blank(),
        panel.grid.minor = element_blank()) 


```
